# Chapter 6 - Longitudinal analysis

Seppo Nyrkkö _Fri Dec 4, 2020_

## Reading the data in.

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Rats diet weighting data in long format
longrats <- read.csv("data/ratslong.csv")

# BPRS patient treatment scores
longbprs <- read.csv("data/bprslong.csv")
```

## i) RATS analysis

```{r}

summary(longrats)
str(longrats)

```
### Quick summary of the long data (Rats):
* Rats have unique IDs.
* Control group category (factorized) is the diet Group
* Time axis is numeric in days
* Longitudinal measurements are weights

```{r}
#  Some ggplots

# Display the data graphically, first by groups.

ggplot(longrats, aes(x = days, y = weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(longrats$weight), max(longrats$weight)))

```

Seems like the grouping is done and handled correctly.
Time goes in the right direction and weight grows as the days pass.

Group 1 doesn't gain weight as much as groups 2 and 3.
An outlier rat in group 2 is a bit heavier but still 
belongs to the correct diet group.


Since the initial weights differ so much we can understand better
the effects of the diets by standardizing the data
```{r}
# Standardizing = zero mean, unit = standard deviance
longrats <- longrats %>%
  group_by(days) %>%
  mutate(stdWeight = (weight - mean(weight)) / sd(weight)) %>%
  ungroup()
str(longrats)

```
This seems to adjust our weights per normal growth.
Again, graphically:
```{r}
ggplot(longrats, aes(x = days, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "std weight")
```

This looks much promising. The control group (1) seems to stay
at the center, and two diet groups (2 and 3) seem to affect the
weight at some direction as the time passes.

Now calculate the group means and errors.
```{r}

# some fancy pipelining 
ratgroups <- longrats %>%
  group_by(Group, days) %>%
  summarise(mean = mean(weight), se = sd(weight) / sqrt(length(weight))) %>%
  ungroup()

summary(ratgroups)
dim(ratgroups)
```

Now we get a graphical analysis how the groups react to the diets
per days run:
```{r}
ggplot(ratgroups,
  aes(x = days, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="2"), width=1) +
  theme(legend.position = c(0.8,0.8,0.8)) +
  scale_y_continuous(name = "mean(weight) +/- std err")
```


To validate our rats, we can do the boxplot with error bars
to visually determine whether
our weight groups are okay or if they contain outlier rats.

```{r}
ratbox <- longrats %>%
  filter(days > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(weight) ) %>%
  ungroup()

```

```{r}
ggplot(ratbox, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=1, size=8, fill = "white", col="red") +
  scale_y_continuous(name = "mean(weight) days from 1 upwards")

```


```{r}
summary(ratbox$Group)
```
Since there are only 4 rats in groups 2 and 3, it's really hard
to say if there are any outliers. We are happy with the one
suspiciously heavy rat in group 2. 

**Let's continue to the hypothesis testing then.**
The first question is whether the mean weight of group 1 differs from others.

```{r}
# Two-sample t-test
ratbox$isGrp1 <- ratbox$Group=="Grp1"
t.test(mean ~ isGrp1, data=ratbox, var.equal=T)

```

This gives the 95% confidence range of group 1 as weighing from 202 to 284 grams.


**Now adding a initial weight from the first day before the diets**

```{r}
# Since we are interested in the growth profile of these rats we take in account
# the day 1 weight

ratinit <- longrats %>%
  filter(days == 1) %>%
  group_by(Group, ID) %>%
  summarise(initialW=mean(weight) ) %>%
  ungroup()

ratbox$initialW=ratinit$initialW

fit <- lm(mean ~ initialW + isGrp1, data=ratbox)
fit

anova(fit)
```
It seems like the group 1 has a 95% confidence to be of a different
growth profile. The visual examination supports this interpretation.



## ii) BPRS analysis


```{r}
str(longbprs)
summary(longbprs)

```
### Quick summary of the long data:
* Subjects have unique subject-IDs (factorized).
* Group category is the treatment id (factorized) 
* Time axis is numeric in weeks
* Measurements are bprs score 


```{r}
glimpse(longbprs)
summary(longbprs)
```
Let's visually analyze the BPRS study groups:
```{r}

ggplot(longbprs, aes(x=weeks, y=bprs, linetype=subject)) +
  geom_line() +
  scale_linetype_manual(values=rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller=label_both) +
  scale_y_continuous(name="BPRS", limits=c(0,100)) +
  theme(legend.position="top")

```

Okay this looks really good. At least the data is read in correctly and
this correlates to what we read in from the original bprs survey data.

We cannot be exactly sure whether the scores go down with the treatment
T1 group, but let's continue with the analysis. The variance in T2 scores
makes it difficult to visually compare the trend during the treatment weeks.

Let's try some regression modeling

```{r}

bprsreg <- lm(bprs ~ weeks + treatment, data=longbprs)
summary(bprsreg)
```

This doesn't show yet any significance of the treatment,
as regards to the group means,
but the weeks undergone seems to be a significant variable.

Since the observations are from repeating individuals,
we would like to introduce some **mixed models** here.
We can try to see whether we allow the **unique subject id** to
count as a *random effect*:


```{r}
library(lme4)

# Random intercept model per subject:

bprs_sub <- lmer(bprs ~ weeks + treatment + (1 | subject), data=longbprs, REML = F)
summary(bprs_sub)

```

```{r}
# Now random intercept and ALSO A random slope model

bprs_subweek <- lmer(bprs ~ weeks + treatment + (weeks | subject), data=longbprs, REML = F)
summary(bprs_subweek)

```

**Notice the higher estimate of treatmentT2** -- This may prove to be significant.

The week-based individual slope has quite a big effect (variance at around 2).

Now let's run an ANOVA test against these two random effects models

```{r}
anova(bprs_sub, bprs_subweek)
```

**We would like to conclude that the model which takes the subject id and the personal slope as a random effect fits the data quite well!**

At 95 percent confidence level we can consider the treatment groups different!

Let's plot the fitted curves to examine this visually

```{r}

# Compute the fitted values 
Fitted <- fitted(bprs_subweek, longbprs)

# Add a new column for the fitted data
bprslongfit <- mutate(longbprs, bprs=Fitted)

# Draw a plot of bprs (fitted and observed)
ggplot(bprslongfit, aes(x=weeks, y=bprs, group=subject)) +
  geom_line(color="blue") +
  geom_line(data=longbprs,color="red") +
  scale_x_continuous(name="weeks") +
  facet_grid(. ~ treatment, labeller=label_both) +
  scale_y_continuous(name="BPRS fitted (blue) vs observed (red)") +
  theme(legend.position="top")
```

This comparison helps understanding the random effects rather well. Also the fitted lines of the observations help seeing the higher BPRS level in the other treatment group, detected by the mixed effects model. 