# Chapter 5 - PCA and MCA
Seppo Nyrkkö _Fri Nov 27, 2020_


## i. reading the human equality data

```{r}

library(ggplot2)
library(GGally)

# read human data, row names in first column


human <- read.csv("data/human.csv",row.names=1)

# examine the structure

str(human)
dim(human)
summary(human)
```

The data has country as row name, ratio of 2nd education M/F, ratio of labor F/M,
expected years of education, exp years of life, GNI of the nation, maternal mortality,
adolescence birth rate and female parliament representation ratio.

Let's see some ggpairs:

```{r}
ggpairs(human)
```

There might be some positive correlation between GNI and education years, and life years.
Maternal mortality and adolescence birth  has inverse correlation with GNI. 

```{r}
hist(human$gni,breaks = seq(0,1,0.2)*max(human$gni),col='blue')
```
The highest GNI per capita rates are only in a few countries. Bear this in mind.


## ii. Some PCA


First, some PCA on the non-standardized data. 

```{r}
pc_human <- prcomp(human)
pc_human
summary(pc_human)
```

See the large differences in deviances.

```{r}
biplot(pc_human)
```

This also gives a "not too clear" biplot. The data seems to be on a single axis: gni.
(Might be caused from the fact that dollars and percentages are of different ranges)

## iii. PCA with standardized data 

We'll continue to standardizing the variables, and do the PCA again.

```{r}
stdpc_human <- prcomp(scale(human))
stdpc_human
```
Exploring some dimensions:
```{r}
barplot(stdpc_human$rotation[,1])
title('PC1: high maternal mortality and adolescent birth rate')
```

```{r}
barplot(stdpc_human$rotation[,2])
title('PC2: high female labour ratio and parliament representation')
```



```{r}
summary(stdpc_human)
```

## iv. Standardized human data biplot

```{r}
biplot(stdpc_human)
```

Now the component arrows are distinguishable. Also the biplot is clerer to read as a "map".

The PC1 component shows the dimension where GNI is low and maternal mortality and adolescent birth rate is high. The PC2 component shows a dimension where female labor is common but also parliament representation rate is higher.




## v. Tea dataset

```{r}
# some libraries
library(FactoMineR)
library(dplyr)
library(tidyr)

```

Let's use the tea dataset which is a survey of tea consumption.
300 individuals answered, how they drink tea (18 Q), what are their product's perception (12 Q) and some personal details (4 Q).

```{r}
data(tea)
str(tea)
```



```{r}
# reduced dataset
tea_segments <- dplyr::select(tea, c("Tea", "relaxing", "effect.on.health", "sugar", "age_Q", "sex"))
str(tea_segments)
dim(tea_segments)
summary(tea_segments)

```

Now let's do Multiple Correspondence Analysis on the reduced tea data

```{r}
# some visualization
gather(tea_segments) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 10))


```

We can see our factors are populated enough in all categories, in order to make some further analysis of our reduced data set.

Now continuing to the MCA (which is kind of categorial PCA):

```{r}
mca <- MCA(tea_segments, graph = FALSE)
# summary of the model
summary(mca)
# visualize MCA
plot(mca, invisible=c("ind"), habillage = "quali")
```

This gives a rather nice plot to understand the correlation between young age group (15-24) and Earl Grey tea, which is the most commonly available blend in tea.

Black tea is common choice for all age groups over 35 years. 

Sugar consumption is more common in the lower age groups, as the taste buds are not yet 'driven in', so to say.

If designing a set of tea flavors, this graph could give a hint to create product packages and images appealing to the targeted market segments.

